```{r,message=FALSE,echo=FALSE}
  html <- TRUE
  library(knitcitations)
  library(poppr)
  bib <- read.bibtex("bibtexlib.bib")
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE, 
                 fig.width = 10, fig.height = 6, cache = TRUE)
  if (html) opts_chunk$set(out.width = "700px", dpi = 300)
  # use this to set knitr options: 
  # http://yihui.name/knitr/options #chunk_options
```
---
title: "6: Locus stats, heterozygosity, HWE"
---

A rigorous population genetic analysis looks closely at the data to assess quality and identify outliers or problems in the data such as erroneous allele calls. This chapter focuses on analysis on a per-locus level. While there are statistics that analyze populations across loci, it is important to analyze each locus independently to make sure that one locus is not introducing
bias into the analysis. 

> **Note:** Many of these statistics are specific to co-dominant data. 

Locus summary statistics
-----

A quick way to assess quality of the data is to determine the number, diversity,
expected heterozygosity, and evenness of the alleles at each locus. As an 
example, we will use data for the fungal-like protist *Phytophthora infestans* from 
`r citep(bib["goss2014irish"])`. First, we'll use the function `locus_table` to get 
all of the statistics mentioned above. For documentation on this function type 
`?locus_table`. Here is a first look at each locus:

```{r, locus_table_stats}
library(poppr)     # Make sure poppr is loaded if you haven't done so already.
data(Pinf)         # P. infestans data set from Mexico and South America
locus_table(Pinf)
```

We can see here that we have widely variable number of alleles per locus and 
that we actually have a single locus that only has two alleles, 
`r {x <- locus_table(Pinf); rownames(x)[x[, 1] == 2]}`. 
This locus also has low diversity, low expected heterozygosity
and is very uneven in allele distribution. This is a sign that this might be a phylogenetically
uninformative locus, where we have two alleles and one is occurring at a minor
frequency. We should explore data analysis with and without this locus. Let's
first see if both of these alleles exist in both populations of this data set.

```{r, locus_table_pops}
locus_table(Pinf, pop = "North America")
locus_table(Pinf, pop = "South America")
```

Phylogenetically uninformative loci
-----

We can see that the South American populations are fixed for this allele, thus 
it would not be a bad idea to remove that locus from downstream analyses. We can
do this using the function `informloci`. This will remove loci that contain less
than a given percentage of divergent individuals (the default is $2/N$, where 
$N$ equals the number of individuals in the data set). 

```{r, informloci}
nLoc(Pinf)  # Let's look at our data set, note how many loci we have.
iPinf <- informloci(Pinf)
nLoc(iPinf) # Note that we have 1 less locus
```

So, how does this affect multi-locus based statistics? We can see immediately 
that it didn't affect the number of multilocus genotypes, let's take a look at
the index of association:

```{r, iatest}
poppr(Pinf)
poppr(iPinf)
```

We can see that it increased ever so slightly for the "North America" and "Total" populations, but not the "South America" population as expected given the fixed alleles at locus P33.

Missing data
----

It is often important to asses the percentage of missing data. The *poppr* function `info_table` will help you visualize missing data so that you can assess how to treat these further using the function `missingno`. For this example, we will use the nancycats data set as it contains a wide variety of possabilities for missing data:

```{r, missing_table, fig.height=6, fig.width = 10, fig.cap = "Plot of missing data"}
data(nancycats)
info_table(nancycats, plot = TRUE)
```

Here we see a few things. The data set has an average of 2.34% missing data
overall. More alarming, perhaps is the fact that population 17 has not been 
genotyped at locus fca45 at all and that locus fca8 shows missing data across
many populations. Many analyses in *poppr* can be performed with missing data in
place as it will be either considered an extra allele in the case of MLG
calculations or will be interpolated to not contribute to the distance measure
used for the index of association. If you want to specifically treat missing data,
you can use the function `missingno` to remove loci or individuals, or replace 
missing data with zeroes or the average values of the locus. 

Removing loci and genotypes
-----

When removing loci or genotypes, you can specify a cutoff representing the percent missing to be removed. The default is `0.05` (5%).

```{r, fig.cap = "Plot of missing data"}
nanloc <- missingno(nancycats, "loci") # Removing loci with greater than 5% missing.
miss <- info_table(nanloc, plot = TRUE, scale = FALSE)
```

> **Advanced Users:** when `scale = TRUE`, the color scale will be set so that 
the warmest color corresponds to the highest value. 

We only removed two loci. If we wanted to make sure we removed everything, we
could set `cutoff = 0`.

```{r}
nanloc <- missingno(nancycats, "loci", cutoff = 0)
miss <- info_table(nanloc, plot = TRUE)
```

Again, removing individuals is also relatively easy:

```{r, fig.cap = "Plot of missing data"}
nanind <- missingno(nancycats, "geno")
miss <- info_table(nanind, plot = TRUE)
nanind <- missingno(nancycats, "geno", cutoff = 0)
miss <- info_table(nanind, plot = TRUE) 
```

`Missingno` removes individuals based on the percent of missing data relative to
the number of loci. Let's remove all individuals with 2 missing loci:

```{r, fig.height=6, fig.width = 10, fig.cap = "Plot of missing data"}
nanind <- missingno(nancycats, "geno", cutoff = 2/nLoc(nancycats))
miss <- info_table(nanind, plot = TRUE) 
```

We only found one individual in population 11.

Replacing missing data
----

Replacement of missing data occurs for each allele over all loci. It will replace all missing data in your data set. There are two options: replacement of
missing data with zeroes, in fact recoding these as another
allelic state, or replacement of missing data with the average allele 
frequency observed. 

Note that the first population in the data set has 20% missing data
at the first locus. Here is the un-replaced data for reference:

```{r}
nan1 <- popsub(nancycats, 1, drop = TRUE) # Dropping alleles not in that population.
nan1[loc = "L1"]@tab
```

```{r}
nanzero <- missingno(nancycats, "zero")
popsub(nanzero, 1, drop = TRUE)[loc = "L1"]@tab
```

The `NA`s have been replaced with zeroes. Now let's look at what happens when we
replace with `"mean"`.

```{r}
nanmean <- missingno(nancycats, "mean")
popsub(nanmean, 1, drop = TRUE)[loc = "L1"]@tab
```

Notice that there are a lot more alleles than there were originally. This is
because the procedure is performed over the entire data set, not by population.
Let's look at what happens if we perform the same routine on the subsetted data.

```{r}
nan1 <- popsub(nancycats, 1, drop = TRUE) # dropping alleles not in this population. 
nan1mean <- missingno(nan1, "mean")
nan1mean[loc = "L1"]@tab
```

Hardy-Weinberg equilibrium
-----

Next, let's determine if our  populations are in Hardy-Weinberg equilibrium. We will again use the nancycats data to test for HWE using genind objects, a dataframe particular to the *adegenet* R package: 

```{r, hwe}
if (!require(genetics)) 
  install.packages("genetics", repos = "http://cran.at.r-project.org")
nanhwe.full <- HWE.test.genind(nancycats) # returns a list
nanhwe.mat  <- HWE.test.genind(nancycats, res.type = "matrix") # returns a matrix
```

The full report (`nanhwe.full`) will give us a list of loci, each containing a list of 
populations, where each population is the result of a $\chi^2$ test. We will analyze data for a few loci to avoid writing a large amount of data to the screen. With the following statements, we can see that at locus1 (`fca8`), we have no good evidence to reject the null hypothesis that population 1
(`P01`) is not in HWE, but we have evidence that population 2 (`P02`) is not in 
HWE:

```{r}
nanhwe.full$fca8$P01
nanhwe.full$fca8$P02
```

Of course, checking all of these is tedious, but we can look at a matrix for all loci by population:

```{r}
nanhwe.mat
```

However, this is still hard to sift through. We will thus define an $\alpha$ value
and set everything above that value to 1 so that we can visually detect candidate 
loci where we might reject the $H_o$ of HWE.

```{r}
alpha  <- 0.05
newmat <- nanhwe.mat
newmat[newmat > alpha] <- 1
```
Now we can create a simple heatmap with `levelplot`.

```{r, fig.height=6, fig.width = 10, fig.cap = "Heatmap showing significant departures from HWE"}
library(lattice)
levelplot(newmat)
```

This simple plot shows us loci in rows and populations in columns. Note, that 
all loci shown in pink are loci suspected for not being in HWE with $p \leq 0.05$. 

References
----------
```{r,echo=FALSE, results='asis', cache = FALSE}
  bibliography(style = "html", bulleted=FALSE)
```