```{r,message = FALSE,echo = FALSE}
  library("knitcitations")
  library("knitr")
  cite_options(citation_format = "pandoc", max.names = 3, style = "html", hyperlink = "to.doc")
  bib <- read.bibtex("bibtexlib.bib")
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE,
                 fig.width = 10, fig.height = 6, cache = TRUE) 
  # use this to set knitr options: 
  # http://yihui.name/knitr/options #chunk_options
```
---
title: '11: Discriminant analysis of principal components (DAPC)'
---

# Introduction

Often we want to infer population structure by determining the number of
clusters (groups) observed without prior knowledge. Several approaches can be
used to infer groups such as for example K-means clustering, Bayesian clustering
using STRUCTURE, and multivariate methods such as Discriminant Analysis of
Principal Components (DAPC) 
`r citep(bib[c("jombart2010discriminant", "pritchard2000inference", "grunwald2011evolution")])`. 
A STRUCTURE-like approach assumes that markers are not linked and that
populations are panmictic `r citep(bib[c("pritchard2000inference")])`. To use
model-free methods K-means clustering based on genetic distance or DAPC are more
convenient approaches for populations that are clonal or partially clonal. Here
we explore DAPC further.

# DAPC analysis of the H3N2 influenza strains

DAPC was pioneered by Jombart and colleagues `r citep(bib[c("jombart2010discriminant")])` 
and can be used to infer the number of clusters of genetically related
individuals. In this multivariate statistical approach variance in the sample is
partitioned into a between-group and within- group component, in an effort to
maximize discrimination between groups. In DAPC, data is first transformed using
a principal components analysis (PCA) and subsequently clusters are identified
using discriminant analysis (DA). This tutorial is based on the
[vignette](http://adegenet.r-forge.r-project.org/files /tutorial-dapc.pdf)
written by Thibaut Jombart. We encourage the user to explore this vignette
further. The vignette can also be opened within R by executing
`adegenetTutorial("dapc")`.

We will use the seasonal influenza dataset H3N2 data containing 1903 isolates
genotyped for 125 SNPs located in the hemagglutinin segment. This dataset as
well as the `dapc()` function is part of the
[*adegenet*](http://adegenet.r-forge.r-project.org) package.

```{r}
# DAPC requires the adegenet package. Let's load this package:
library("adegenet")
data(H3N2) # load the H3N2 influenza data. Type ?H3N2 for more info.
pop(H3N2) <- H3N2$other$epid
dapc.H3N2 <- dapc(H3N2, var.contrib = TRUE, scale = FALSE, n.pca = 30, n.da = nPop(H3N2) - 1)
scatter(dapc.H3N2, cell = 0, pch = 18:23, cstar = 0, mstree = TRUE, lwd = 2, lty = 2)
```

The `dapc()` arguments we used refer to:

- the dataset H3N2
- `var.contrib` this is set to `TRUE`, meaning that we want to retain the
  variables contributing to the analysis in our output. We will use this later
  to see which loci are responsible for separating populations.
- `center` this is set to `FALSE`, indicating that we do not want the data to be
  rescaled so the mean = 0.
- `n.pca` is the number of axes retained in the Principal Component Analysis
  (PCA). By default, it is set to `NULL`.
- `n.da` is the  number of axes retained in the Discriminant Analysis (DA). By
  default, it is set to `NULL`.

> It is important to set `n.pca = NULL` when you analyze your data because the
> number of principal components retained has a large effect on the outcome of
> the data. See the section below for a statistical method called cross-
> validation as an aid for choosing `n.pca`

The `scatter()` function is part of the *ade4* package and plots results of a
DAPC analysis.

As you can see, each year between 2001 to 2005 is a cluster of H3N2 strains
separated by axis 1. In contrast, axis 2 separates the strains observed in the
2006 cluster from the clusters observed during 2001-5, indicating that the
strains observed in 2006 are genetically distinct.

Next, let's assess if there are alleles that most differentiate the 2006 cluster
from those in other years.

```{r}
set.seed(4)
contrib <- loadingplot(dapc.H3N2$var.contr, axis = 2, thres = 0.07, lab.jitter = 1)
```

It looks like SNPs at position 384 and 906 are involved. Let's check this
further by looking at allele frequencies by year:


```{r}
temp    <- seploc(H3N2)       # seploc {adegenet} creates a list of individual loci.
snp906  <- tab(temp[["906"]]) # tab {adegenet} returns a matrix of genotypes
snp399  <- tab(temp[["399"]])

# The following two commands find the average allele frequencies per population
(freq906 <- apply(snp906, 2, function(e) tapply(e, pop(H3N2), mean, na.rm = TRUE)))
(freq399 <- apply(snp399, 2, function(e) tapply(e, pop(H3N2), mean, na.rm = TRUE)))
```

Note that a new allele appeared in 2005 for SNP locus 906 and 2004 for locus 399
separating populations along axis 2.


```{r}
# First, set the plotting parameters
# mfrow = number of columns, rows
# mar   = plot margin size
# las   = axis label style (3: always vertical)
par(mfrow = c(1, 2), mar = c(5, 4, 4, 0) + 0.1, las = 3)

matplot(freq906,  pch = c("c", "t"), type = "b", 
        xlab = "year", ylab = "allele frequency", main = "SNP # 906",
        xaxt = "n", cex = 1.5)
axis(side = 1, at = 1:6, lab = 2001:2006)

matplot(freq399, pch = c("c", "t"), type = "b", 
        xlab = "year", ylab = "allele frequency", main = "SNP #399",
        xaxt = "n", cex = 1.5)
axis(side = 1, at = 1:6, lab = 2001:2006)

# Now we reset the plotting parameters to default
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1, las = 0)
```

This plot nicely illustrates the effect of mutation, followed by selection or
drift in the seasonal H3N2 influenza virus.

# Cross validation: DAPC analysis of *Phytophthora ramorum* from forests and nurseries

Above we showed a nice example of a story that shows how two loci can
drastically influence an epidemic. Next we will spend some time establishing what the appropriate
number of principal components (PC) is for the analysis. It is important
to carefully choose the correct number of PCs so as to include most sources of
variation explained by an appropriate number of PCs retained. One way of
ensuring that you have selected the correct number of PCs is to do cross-validation. 
This is a procedure in which you leave out a certain percentage of
your data, run DAPC, and then see if the data that was left out is correctly
placed into the population.

Since the H3N2 data set is quite large, we will use *Phytophthora ramorum* data
from nurseries in California and Oregon `r citep(bib["goss2009population"])` and
forests in Curry County, Oregon `r citep(bib['kamvar2015spatial'])`. These data
represent the Sudden Oak Death epidemic in Curry County, OR from 2001-2014
separated into different watershed regions. In `r citet(bib["kamvar2015spatial"])`,
the "Hunter Creek (HunterCr)" population was shown to be the result of a second
introduction, likely from nurseries. Part of the evidence to support this
conclusion came from DAPC results. Here, we will recreate the process of cross
validation and reporting.

If we run the function `xvalDapc()` with default parameters, it will run 30
replicates of cross-validation for a number of PCs less than the total number
of alleles in the data. This is a good way to get an idea of where to focus
more intense cross-validation runs:

> By default `xvalDapc()` needs two parameters: 1. The genotype matrix, 2. The
> population factors.

```{r}
library("poppr")
data("Pram", package = "poppr")
Pram
set.seed(999)
pramx <- xvalDapc(tab(Pram, NA.method = "mean"), pop(Pram))
```

You can see that we have a peak around 15 PC. From here, we can narrow the search
by specifying the number of PC to try with `n.pca` and centering it around 15,
and doing 1000 replicates each (Note, this will take a long time).

> Windows users: change parallel to "snow".

```{r, Pramhide}
set.seed(999)
system.time(pramx <- xvalDapc(tab(Pram, NA.method = "mean"), pop(Pram), 
                             n.pca = 10:20, n.rep = 1000, 
                             parallel = "multicore", ncpus = 4L))
```

We can see that it's basically a flat line all the way. If we take a look at
the object, we see that 16 PCs give us the highest percent of correctly predicted
subsamples with the lowest error.

```{r }
names(pramx) # The first element are all the samples
pramx[-1]
```

We also have a DAPC object that we can plot comparable to figure 4
in `r citet(bib["kamvar2015spatial"])`:

```{r}
scatter(pramx$DAPC, col = other(Pram)$comparePal, cex = 2, legend = TRUE, 
        clabel = FALSE, posi.leg = "bottomleft", scree.pca = TRUE, 
        posi.pca = "topleft", cleg = 0.75, xax = 1, yax = 2, inset.solid = 1)
```

We can see that this shows the clear separation of Hunter Creek from the rest
of the epidemic, providing evidence that this population arose from a separate introduction.

# Conclusions

DAPC is a wonderful tool for exploring structure of populations based on PCA and
DA without making assumptions of panmixia. Thus, this technique provides a
robust alternative to Bayesian clustering methods like STRUCTURE 
`r citep(bib[c("pritchard2000inference")])` that should not be used for clonal or
partially clonal populations.

DAPC analysis is inherently interactive and cannot be scripted *a priori*.
Please refer to the 
[vignette](https://github.com/thibautjombart/adegenet/blob/master/tutorials/tutorial-dapc.pdf) 
written by Thibaut Jombart for a more interactive analysis.

# References
