---
title: "genlight_dist.Rmd"
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Distance matrices

Let's create a pairwise genetic distance matrix for individuals or populations (i.e., groups of individuals).

> **Note**: There isn't actually a function that creates distance matrices from genlight objects in adegenet.
Instead, the authors of adegenet created an `as.matrix()` function that converts a genlight object to a matrix. This is clever because the function `dist()` in the package `stats` tries to convert whatever object it is given into a matrix.
The result is that when you call `dist()` on a genlight object it uses the `dist()` function to create a distance matrix.
The reason this is clever is because it uses pre-existing code.
The downside is that because there is no function to specifically create distance matrices from genlight objects in adegenet, there is no documentation in genlight for how this is done.
And because the author of `dist()` never anticipated it could be used on genlight objects, there is no documentation for it there either.

To summarize, we can create a distance matrix from a genlight object using `dist()`:

```{r dist, eval=FALSE}
x.dist <- dist(x)
```

Note, that we have not specified what the variable x is. 
We can find documentation for this function with `?dist`.

There are also functions to create distance matrices from genlight objects that exist in other packages.
The function `bitwise.dist()` in the package *poppr* is an example.
We can find documentation for this function with `?poppr::bitwise.dist`.
Again, you need to know where to look for this information or you may not find it.
We can use this function as follows.


```{r bitwise.dist, eval=FALSE}
x.dist <- poppr::bitwise.dist(x)
```

Note, that the variable x has not yet been specified. 
Lastly, because you can use `as.matrix()` on your genlight object, and most distance algorithms can use this matrix as input, you can use this as an intermediate step to create a matrix from your genlight object and pass it to your distance algorithm of choice.
Options include *ade4*, `vegdist()` in *vegan*, or `daisy()` in *cluster.*
Note that it is up to you to determine which distance metric is best for your particular analysis.
A number of options therefore exist for creating distance matrices from genlight objects.
