---
title: "genlight_dist.Rmd"
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Distance matrices

Let's create a pairwise genetic distance matrix for individuals or populations (i.e., groups of individuals).

To summarize, we can create a distance matrix from a genlight object using `dist()`:

```{r dist, eval=FALSE}
x.dist <- dist(x)
```

Note, that we have not specified what the variable x is. 
We can find documentation for this function with `?dist`.

There are also functions to create distance matrices from genlight objects that exist in other packages.
The function `bitwise.dist()` in the package *poppr* is an example.
We can find documentation for this function with `?poppr::bitwise.dist`.
Again, you need to know where to look for this information or you may not find it.
We can use this function as follows.


```{r bitwise.dist, eval=FALSE}
x.dist <- poppr::bitwise.dist(x)
```

Note, that the variable x has not yet been specified. 
Lastly, because you can use `as.matrix()` on your genlight object, and most distance algorithms can use this matrix as input, you can use this as an intermediate step to create a matrix from your genlight object and pass it to your distance algorithm of choice.
Options include *ade4*, `vegdist()` in *vegan*, or `daisy()` in *cluster.*
Note that it is up to you to determine which distance metric is best for your particular analysis.
A number of options therefore exist for creating distance matrices from genlight objects.
