


### Discriminant analysis of principal components (DAPC)

- Reorganize DAPC plots
- Explain cross-validation plots
- Explain in detail what a DAPC plot show

Another way of determining if the geographic populations are clustered is to use a discriminant analysis of principal components (DAPC). 
DAPC is a multivariate statistical approach that uses groups defined a priori to maximize among group variance and minimize within group variance.
DAPC is explained in depth in the [DAPC chapter](DAPC.html) on Part II of this tutorial or on the DAPC [adegenet vignette](http://adegenet.r-forge.r-project.org/files%20/tutorial-dapc.pdf). 


The DAPC function is designed to work with genlight object **with popluations defined a priori**.
For this excercise we will use the genlight object with all the SNP's. 


First, we will remove the *P. mirabilis* sample to allow us to focus on how the *P. infestans* samples group.


```{r}
set.seed(1234)
x <- popsub(x, blacklist = "Pmir")
dapc.x <- dapc(x, pop = pop(x), n.pca=10, n.da = 2)
scatter(dapc.x, scree.da=FALSE,  bg="white", pch=17:22, cstar=0, col=cols)
scatter(dapc.x, col = cols, cex = 2, legend = TRUE, clabel = F, posi.leg = "bottomleft", scree.pca = TRUE,
        posi.pca = "topleft", cleg = 0.75)
```


This figure doesn't allow us to see any particular pattern, that why we use cross-validation. 
Very generally, cross-validation will objectivelly optimize the best trade-off point between retaining too many or too few PC's for our model.  The [DAPC chapter](DAPC.html) on Part II of this tutorial explains why cross-validation is important. We need a genclone/genind object for this: the `x.gc.popsub` object. This object is a genclone object without the *P. mirabilis* sample:


```{r}
mat <- as.matrix(x)
grp <- pop(x)

xval <- xvalDapc(mat, grp, n.rep = 30, n.pca = c(1:10))
```

The cross-validation shows that 5 PC is the ideal point of correctly predicted subsamples with the lowest error. we can plot the cross-validation DAPC results:

```{r}
set.seed(12345)
scatter(xval$DAPC, col = cols, cex = 2, legend = TRUE, clabel = TRUE, posi.leg = "bottomleft", scree.pca = TRUE,
        posi.pca = "topleft", cleg = 0.75)
```

The DAPC plot shows what we've found using all the other methods until now: There is no population aggrupation between the geographic locations of samples of *P. infestans*

> Javier's Opinion: This excercise is very long without all the other topics. We should discuss what we want for the workshop vs what we want for the book.


