```{r,message=FALSE,echo=FALSE}
  library("knitcitations")
  library("knitr")
  cite_options(citation_format = "pandoc", max.names = 3, style = "html", hyperlink = "to.doc")
  bib <- read.bibtex("bibtexlib.bib")
  opts_chunk$set(tidy = FALSE)
```
---
title: "A3: Introduction to R"
---

This appendix introduces basic use of the [R](http://www.r-project.org) statistical and computer language
and the [Rstudio](http://www.rstudio.com/ide/) graphics user interface. We assume you installed the needed resources mentioned in chapter 2 on getting ready to use R.

After installing both R and R studio, please proceed to open R studio. R studio should look something like:

![Rstudio](images/Rstudio.tiff)

The top left panel is a text-box where you can write a script. The bottom left panel is the console (or command line) where you directly execute commands and R writes results to the console; for example, type in `x <- 1` hit return and then type `x` and hit return again and watch the console:

```{r}
x <- 1
x
```

You just assigned the value 1 to the variable `x` with the assignment operator $<-$. The second command asks R to show the value assigned to R. Finally, the top right panel shows variables loaded into the environment and the bottom right panel is useful for loading files, plotting graphs, installing packages and getting help. Feel free to explore all the various tabs or elements of each panel. 

Assignment, data types and operations
---

Let's learn a few simple facts about R:

R uses # to add comments to code:
```{r}
# Add 1 + 1 and write to console
1+1
```

This is very useful when developing reusable code or when you want to remember a few months later what you actually did.

R is case sensitive so that `x = 5` is not the same as  `X = x + 1`. Try typing this into your console and look at `x` and `X`, respectively:
```{r}
x = 5
X = x+1
x
X
```

Next, let's assign numbers to a vector:
```{r}
x <- c(1,2,3,4,5)
x
```

You can see several important traits of R. The `c()` command (which stands for combine) can be used to make a list (e.g., a vector) of data. We can overwrite the variable `x` with new content and we can assign another datatype (e.g., a single number becomes a vector). To compound things further you can even change the data type to strings overwriting the previous content:
```{r}
a <- c("Paris", "Tokyo", "Seattle")
a
```
Note that we use quotes to denote the string data type containing text. 

To access individual elements of a vector, in this case the second element, we execute:

```{r}
a[2]
```

Console tricks: Code completion & command history
---

R studio has a very useful feature called code completion using the **Tab** key which can complete the full name of an object. For example type `hel` and hit **Tab** and you will see several functions pop up and you can select `help()`. 

This also work inside a function to find function arguments. Type `help(` and hit *Tab* to select arguments for the help function.

R studio records your command history. Thus you can scroll up or down the history of executed commands using the *Up* or *Down* arrows. Make sure your cursor is in the console and try to re-execute previous commands.

Useful R functions
---

Other useful commands providing information on your system include:

```{r, eval=FALSE}
help(options)                                 # learn about available options
installed.packages()[,c("Package","Version")] # list packages & versions installed
sessionInfo()    						  # displays packages currently loaded
data(<name>, package = "package")             # load a specific data set from a specific package
```

To quit R you can either use the `Rstudio > Quit` pull-down menu command or execute `⌘ + Q` (OS X) or `CTRL + Q` (PC). 

Getting help
---

One way that R shines above other languages for analysis is the fact that
R packages in CRAN are all documented. Help files are written in HTML and give
the user a brief overview of:

 - The purpose of a function
 - The parameters it takes
 - The output it yields
 - Some examples demonstrating its usage. 

To see all of the help topics in a package, you can simply type:

```{r, eval = FALSE}
help(package = "poppr") # Get help for a package.
help(amova)             # Get help for the amova function.
?amova                  # same as above.
??multilocus            # Search for functions that have the keyword multilocus.
```
Some packages include vignettes that can have different formats such as being
introductions, tutorials, or reference cards in PDF format. You can look at a
list of vignettes in all packages by typing:

```{r, eval=FALSE}
browseVignettes()                     # see vignettes from all packages
browseVignettes(package = 'poppr')    # see vignettes from a specific package.
```

and to look at a specific vignette you can type:

```{r, eval=FALSE}
vignette('poppr_manual')
```

Next, consider browsing the R tutorial chapter if you are not yet familiar with
R and Rstudio. Otherwise, you are now ready to think about formatting and
loading population genetic data into R.


Useful resources
---

*Introductory*

 - [Swirl](http://swirlstats.com/) is a very well thought out R package that teaches you interactively.
 - [Code School Try R](https://www.codeschool.com/courses/try-r) is a nice interactive tutorial. 
 - [Quick R](http://www.statmethods.net/interface/help.html)
 - [R reference card](http://cran.r-project.org/doc/contrib/Short-refcard.pdf)
 - A very nice, short [introduction to R](http://ateucher.github.io/rcourse_site/index.html)

*Advanced*

 - [Advanced R](http://adv-r.had.co.nz/) by Hadley Wickham

*Books*:

 - [R in a Nutshell](http://oreilly.com/catalog/9780596801717) 
 - [R cookbook]() is a nice quick reference and tutorial for general R use.
 - [ggplot2: Elegant Graphics for Data Analysis]() is a useful, bu dated reference if you want to customize graphs for publication. 

<!------------>