---
title: "Missing data"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = 'center')
knitr::opts_chunk$set(fig.width = 12)
```


```{r, results='hide', echo = FALSE}
suppressPackageStartupMessages( library(vcfR) )
#vcf <- read.vcfR('TASSEL_GBS0077.vcf.gz')
data('vcfR_example')
```


As the size of our dataset grow, in terms of samples and variants, the size of our data matrix grows.
As the size of our data matrix grows, it also increases the opportunity to have missing data.
Also, some of our quality filtering steps increased the degree of missingness in our data matrix by setting values that we determined to be of unusual quality to NA.
One way of managing missing data is to use **imputation**, a set of methods that attempts to infer what the most likely genotype should be and replaces the missing genotype with the interpolated genotype.
However, if your data has a large degree of missingness you may want to attempt to mitigate missingness instead of interpolation
Or you may want to implement a mitigation step prior to interpolation in the hope that this will improve the performance of the interpolation.
Missing data can frequently be due to samples (columns) or variants (rows) of low quality.
Here we demonstrate how to identify samples and variants in the data set that have a high degree of missingness.
In another section we'll show how to omit them.


## Overall missingness


A first perspective on how complete our dataset is can be provided by the `show` method for the vcfR object.
When you invoke the name of an object with no arguments it invokes the show method.


```{r}
vcf
```


The show method for an object typically reports a summary of what is contained in the object.
Here we see the number of samples and variants in our data.
We also see a report of what the percentage of missing data is in our object.
In the context of vcfR this is the proportion of variants scored as NA.
Note that if a variant includes some data associated with a missing genotype it will not be recognized as missing.
For example, a missing genotype could be associated with a depth information as follows.


```{r, eval=FALSE}
GT:DP ./.:1
```


Because this variant does include some data it will not be recognized as missing until the genotypes are extracted and queried for missingness.
This means that the degree of missingness reported by the `show` method may be an under representation.
It does provide an easily accessed first perspective on the proportion of missing data.
To determine the cause of this missing data (e.g., are there particular samples or variants of poor quality) we will look further.


## Quantifying missingness, one sample


To quantify missingness for a single sample we can use the function `is.na()`.
This function uses a vector as an argument and returns a logical vector (TRUE and FALSE) indicating which values are NA.
If we remind ourselves that TRUEs and FALSEs are numerically encoded as 1 and 0 it reminds us we can take a sum of this logical vector to determine the degree of missingness.


```{r}
as.numeric(c(FALSE, TRUE))
sum(as.numeric(c(FALSE, TRUE)))
```


Now that we've reminded ourselves of how to count missing values we can apply this knowledge to query one of our samples.
We'll extract a matrix of variant depths (DP) from the VCF data.
We can remind ourselves of what DP is in this VCF data by using the `queryMETA()` function.
(Recall that we have already read in our VCF data to create a vcfR object in the section 'Reading VCF data'.)


```{r}
queryMETA(vcf, "DP")
```


Here we'll want the first definition (FORMAT=ID=DP).
Now we can extract a matrix of depths and query the first sample for missingness.


```{r}
dp <- extract.gt(vcf, element = "DP", as.numeric=TRUE)
sum(is.na(dp[,1]))
```


This reports the number of missing variants in the first sample.
We could similarly count the number of missing samples from a variant by accessing a row instead of a column.
We could also convert this to a percentage by using `length()` to determine the total number of values in either the column or row and use this as a denominator.


## Quantifying missingness, all samples


This is illustrative of what we can accomplish for a single sample or variant.
We typically have many samples an tens of thousand (or more) variants.
We can extend the functionality of the above example to many columns or rows by using the `apply()` function.
See the section on [apply](https://stat.ethz.ch/R-manual/R-devel/library/base/html/apply.html) if you are unfamiliar with this function.
Because we will be summarizing many samples we will use a barplot to visualize the results as opposed to trying to scrutinize the numerical information.


```{r}
myMiss <- apply(dp, MARGIN = 2, function(x){ sum(is.na(x)) })
myMiss <- myMiss/nrow(vcf)

library(RColorBrewer)
palette(brewer.pal(n=12, name = 'Set3'))

par(mar = c(12,4,4,2))
barplot(myMiss, las = 2, col = 1:12)
title(ylab = "Missingness (%)")
par(mar = c(5,4,4,2))
```


This allows us to visualize the degree of missingness on a per sample basis.
We see that the sample P7272 has a particularly high amount of missing genotypes.
This is because this sample is a different species than the reference it was mapped to.
One decision could be to omit this sample.
However, if the goal is to make comparisons among these species we may instead search for variants that are present in both taxa.


We can do something similar to query the variants (rows) for missingness.
However, when we have a large number of variants we wouldn't want to visualize this with a barchart.
It would require a barchart with 60 thousand bars.
Instead of using a barchart we'll use a histogram.



```{r, fig.width=5}
myMiss <- apply(dp, MARGIN = 1, function(x){ sum(is.na(x)) })
myMiss <- myMiss/ncol(vcf@gt[,-1])

hist(myMiss, col = "#8DD3C7", xlab = "Missingness (%)", main = "")
```


We've now seen how we can create summaries of our data matrix over both rows and columns.
Once we have this knowledge in hand we may develop a plan for managing this data.

