```{r,message=FALSE,echo=FALSE}
  library("knitcitations")
  library("knitr")
  cite_options(citation_format = "pandoc", max.names = 3, style = "html", hyperlink = "to.doc")
  bib <- read.bibtex("bibtexlib.bib")
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE, cache = TRUE) 
  # use this to set knitr options: 
  # http://yihui.name/knitr/options #chunk_options
```
---
title: "5: Populations strata: subsetting and clone correction"
---

Populations are best sampled hierarchically on a range of scales from
subpopulations (e.g. fields, valleys, ranges) to regions (e.g. valleys, states,
countries or continents) or across time (years or decades). This approach is
useful because population structure and evolutionary processes may not be
discernible *a priori*. Most of the times we do not know if population are
differentiated spatially or temporaly.  Thus, a combination
of targeted local sampling with sampling over larger spatial or temporal scales
is necessary to detect population structure over different scales, without using
intense sampling throughout the entire range.

The methods implemented in *poppr* allow specification of which strata you
want to analyze.  This is a rapid way of working with subsets of your data
without having to perform any data manipulation or changing the input file.  In
this tutorial, we will show you how to define the hierarchical structure of your
data and  how to specify specific levels that you might want to analyze.

Data used in this example
----
For this example, we will use  the `monpop` data set 
`r citep(bib[c("everhart2014finescale")])`. This microsatellite data consists
of 13 loci for 694 individuals of the haploid fungal pathogen *Monilinia
fructicola* that infects peach flowers and fruits in commercial orchards.  The
monpop population came from four trees within a single orchard (trees 26, 45,
and 7). Each tree was sampled in 2009, 2010, and/or 2011.  Additionally, each
sample was noted as to whether it came from a blossom or a fruit. This example
data set is included with the *poppr* package.

Working with hierarchical data
----

The steps for working with hierarchical data include:

1.  Import data set with samples labeled according to strata
2.  Define the stratifications for the data
3.  Setting the hierarchical level(s) that you want to have analyzed

Importing data labeled according to a strata
----
The easiest way to work with hierarchical data is to label each sample using an
underscore "\_" to separate each hierarchical level.  This was already done for
the monpop data, where each sample was coded hierarchically by tree, year, and
symptom in the following format: "Tree\_Year\_Symptom".

Let's load the hierarchically labeled example data:

```{r}
library("poppr")
data(monpop)
monpop
```


**Genotype information** shows us that the data contains 264 multilocus
genotypes among 694 haploid individuals with 13 loci. **Population information**
has two items, the stratifications and the populations defined.  You can
think of stratifications as the index names for each of the hierarchical
levels within your data (so for our data it should be Tree, Year, and Symptom).
By default, however, no stratifications are defined and so this is "Pop",
which is the entire dataset of 694 individuals.  Because we labeled each sample
according to hierarchical level, populations defined shows us our data has 12
groups defined: `r monpop$pop.names`.

Assigning stratifications
----
We imported the data that has three stratifications "Tree\_Year\_Symptom".
In order to analyze our data according to any combination of those three
stratifications, we need to tell *poppr* that the 12 groups should be split
by tree, year, and/or symptom.  Thus, the first step is to split our data
according to strata so that we can access each of the three hierarchical
levels in the data.  The `splitStrata` command is used to index the three
stratifications:


```{r}
splitStrata(monpop) <- ~Tree/Year/Symptom
monpop
```

> The `splitStrata` command only needs to be run once: right after you import
> your data. 

After splitting the data populations are specified by hierarchical level: "Tree
Year Symptom".

Next, we analyze the data according to Tree and Year:

```{r}
setPop(monpop) <- ~Tree/Year
monpop
```

To analyze the data according to Symptom:

````{r}
setPop(monpop) <- ~Symptom
monpop
````

Order of the levels that you define is important, so if we wanted to define the
symptoms according to tree, we would use the following:

````{r}
setPop(monpop) <- ~Symptom/Tree
monpop
````

Now that we have laid out the basics of manipulating data by strata,
we will now apply strata for clone correction.

Clone correction
----

When dealing with clonal populations, analyses are typically conducted with and
without clone correction. Clone correction is a method of censoring a data set
such that only one individual per MLG is represented per population
`r citep(bib[c('milgroom1996recombination', 'grunwald2003', 'grunwald2006hierarchical')])`. 
This technique is commonly used with the index of association
and genotypic diversity measures since clone corrected populations approximate behavior of sexual populations. Since we want to only observe unique genotypes per population,
clone correction requires specification of the stratifications at which
clones should be censored. This section will show how to clone correct at a
specific hierarchical level and also compare the results with uncorrected data.

> Question: Will allelic diversity increase or decrease with clone-censored data?

Using `monpop` as an example, if we wanted to know the diversity of alleles
within each tree per year, how should we go about correcting for the clones? We
use the function `clonecorrect` specifying the "Tree/Year" strata:

```{r, clonecorrection}
mcc_TY <- clonecorrect(monpop, strata= ~Tree/Year, keep = 1:2)
mcc_TY
```

Notice that the number of samples reduced from `r nInd(monpop)` to 
`r nInd(mcc_TY)`, but is still more than the number of MLGs. This indicates that
there are duplicated genotypes that cross trees and years, but that's okay
because of our definition of a clone-corrected data set as having one
representative genotype *per population*. Before we continue, we should set the
original data to the same strata:

```{r}
setPop(monpop) <- ~Tree/Year
```

Now we can compare the diversity of alleles at each locus using Simpson's index
($1-D$) as implemented in the function `locus_table` (Detailed in our 
[chapter on locus based statistics](Locus_Stats.html)). We will do this in three
steps:

1. Calculate diversity of the clone corrected data
2. Calculate diversity of the uncorrected data
3. Take the difference of **step 1** from **step 2**.

```{r, simpdiv}
cc <- locus_table(mcc_TY, info = FALSE)
mp <- locus_table(monpop, info = FALSE)
mp - cc
```

Let's plot $1-D$ so that we can see it better:

```{r, simpdivplot}
locus_diff <- mp - cc

# Note that I need to select the column containing Simpson's Index. That's
# labeled as "1-D".
barplot(locus_diff[, "1-D"], ylab = "Change in Simpson's Index", xlab = "Locus",
        main = "Comparison of clone-corrected vs. uncorrected data")
```

We can see quite a difference in some loci after clone correcting based on 
tree in the overall data set showing that, while some loci show a decrease,
many loci show an increase in allelic diversity after clone-correction. 

### Advanced R: writing your own functions

Of course, we still want to analyze each tree/year combination separately.
Instead of typing those commands above for each combination, we can write a
function to do it for us.

A function can be thought of as a set of instructions that tells the computer
what to do. You have been using funcitons before such as `poppr()`. A function is written  like this:

```{r, eval = FALSE}
fun_name <- function(arguments){
  # instructions
}
```

`fun_name` is the name of the function, the R command `function` tells R that 
you are about to write a function. Arguments are variables that you will pass
to the function and the curly braces (`{` and `}`) provide the area where you write 
all the steps of the function. Next, we will go through an example of writing 
a function. 

#### Example: harmonic mean

Let's write a function that will take the harmonic mean
of a vector of numbers. First, we'll see how to do this calculation step by step before we write a function.

> The harmonic mean has been utilized to estimate effective population size of a
population that has undergone drastic changes in population size 
`r citep(bib["nielsen2013"])`. 

```{r}
psize <- c(10, 100, 50, 80, 20, 500)
# Arithmetic mean
mean(psize)
# Harmonic mean
rmean <- mean(1/psize)
1/rmean
```

Notice that there are two steps to calculating the harmonic mean. 
If we wanted to calculate the harmonic mean of a lot of different vectors,
this would get really repetitive. A function is a way of encapsulating a lot
of instructions to reduce the number of steps (and potential errors) you would
need to take in your analysis. Below, we will write a function to calculate
the harmonic mean using the steps above. 

> Note here that the *argument* is **x** and it is a placeholder for any value
you put into the function.

```{r}
hmean <- function(x){
  rmean <- mean(1/x)
  1/rmean
}
mean(psize)
hmean(psize)
```

#### Writing a function for comparing clone corrected and uncorrected data

We will simply take the steps from above and turn them into a function to 
calculate the difference in Simpson's diversity for a given population. In order
to do that, we will need to compute three steps:

1. the name of the population
2. the clone-corrected data set
3. the uncorrected data set

From there, we can construct the function.

```{r}
plot_simp_diff <- function(pop_name, clone_corrected, un_corrected){

  # Step 1: calculate diversity for clone-corrected data
  cc <- locus_table(clone_corrected, pop = pop_name, info = FALSE)

  # Step 2: calculate diversity for uncorrected data
  uc <- locus_table(un_corrected, pop = pop_name, info = FALSE)

  # Step 3: Take the difference
  res <- uc - cc

  # Step 4: Plot Simpson's index.
  barplot(res[, "1-D"], main = pop_name, ylab = "Change in Simpson's Index", xlab = "Locus")
}
```

Now that we have our function written, let's gather the population names and
construct a *for loop* that will analyze each one. 

```{r}
par(mfrow = c(2, 3)) # Set up the graphics to have two rows and three columns

for (i in monpop@pop.names){
  plot_simp_diff(i, mcc_TY, monpop)
}

par(mfrow = c(1, 1))
```

These barplots show the difference in Simpson's index of original minus clone
corrected data for each population per locus. We can see that  allelic diversity
generally is lower in the total data set (containing some repeated MLGs)
relative to clone corrected data.

Conclusions
----

This was a brief introduction to the easiest way to create stratifications
and apply them in *poppr* to more rapidly analyze your data. By indexing the
stratifications of your data, you can set the hierarchical level(s) you want
to have analyzed in a single command.  This approach avoids having to
create new sub-sets of the data for each analysis and simultaneously reduces the
chance of error when manipulating data sets by hand.

References
----------

<!-------->
