---
title: "Preparation for the Workshop"
output:
  html_document:
    toc: true
    toc_depth: 2
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Population genomics workshop

### ICPP, Boston, Massachusetts, 2018

Previous versions of this material were also presented at:    
* APS, San Antonio, Texas, 2017

For workshops, we subset the information available elsewhere in this book to the pages in this section specifically for the workshop.
This provides a convenient way to reduce the contents of a book to the essential portions we feel can be reasonably covered in a several hour long workshop.
However, the cost of brevity is a lack of detail.
While this section may be used as an entry point, learning about genomic analyses in R, details may be found elsewhere in the book.


## Before you arrive please make sure you install the following required materials
 

1. __Install R__ 
Instructions on how to install R can be found at the [R homepage](https://www.r-project.org/).
Please make sure you have the most current version of R (3.4.4, Someone to lean on).
2. __Install RStudio__    
Instructions on how to install the RStudio integrated development environment (IDE) can be found at the [RStudio download site](https://www.rstudio.com/products/rstudio/download/).
Choose the 'RStudio Desktop Open Source License' (free) version.
3. __Install R packages__    
You can install the R packages we will be using during the workshop by starting R and copying and pasting the below command into the R console.


```{r, echo = FALSE, eval = FALSE}
myPackages <- system("grep 'library(' *.Rmd", intern = TRUE)
myPackages <- myPackages[grep("library\\(\\)", myPackages, invert = TRUE)]
myPackages <- sub("[[:print:]]*library\\([\'\"]{0,1}", "", myPackages)
myPackages <- sub("[\"\']","", myPackages)
myPackages <- sub("#[[:print:]]*","", myPackages)
myPackages <- sub("[[:space:]]+","", myPackages)
myPackages <- sub("\\)+", "", myPackages)
myPackages <- unique(myPackages)
myPackages <- grep("*.Rmd, intern = TRUE", myPackages, invert = TRUE, value = TRUE)
myPackages <- grep("knitcitations", myPackages, invert = TRUE, value = TRUE)
myPackages <- sort(myPackages)
dput(myPackages)
```


```{r, eval = FALSE}
install.packages(c("adegenet", "ape", "cowplot", "devtools", "dplyr", "ggplot2", "hierfstat", "igraph", "knitr", "lattice", "magrittr", "mmod", "pegas", "pinfsc50", "poppr", "RColorBrewer", "reshape2", "treemap", "vcfR", "vegan", "viridisLite"))
```


4. __Create an RStudio "Project"__    
Open RStudio and under the "File" dropdown menu select "New Project...".
A "New Project" window should pop up where you should select the option to create your project in a "New Directory".
Select the "New Project" option.
You will be asked where on your filesystem you'd like your project and what to name your project.
We'll name our project "APS_Workshop".
This should create a directory for your project that will contain a file named "APS_Workshop.Rproj".
When you double click this file it should spawn a new session of RStudio and will use that directory as your [working directory](https://en.wikipedia.org/wiki/Working_directory).


5. __Download the example datasets __  
We have placed example data files at the OSF site for [Population genomics in R workshop](https://osf.io/ytd6k/).
You can download them by copying and pasting the below code into your R console.


```{r,eval = FALSE}
download.file("https://osf.io/td5sx/download", destfile = "pinfsc50_filtered.vcf.gz")
download.file("https://osf.io/kaq5p/download", destfile = "population_data.gbs.txt")
download.file("https://osf.io/c6h4t/download", destfile = "prubi_gbs.vcf.gz")
```


You should be able to validate that the files have been downloaded with the below code.


```{r}
list.files(pattern = ".vcf.gz$|.gbs.txt")
```


6. __Test that your system is ready:__    
You should now be able to validate that the resources you need are ready.
Change directory to `APS_Workshop` and copy and paste the below command into the R console.


```{r, eval=FALSE}
if(require(devtools)){
  devtools::source_gist("01a3d8efb21e0a4dbac0735270d147af", filename="apstest.R")       
} else {
  print("Please install the package devtools.")
  print("Use: install.packages('devtools')")
}
```


This should output some tests to the console.
It should also generate a report file called `apstest.txt`.
The report file should look exactly like [this file](https://github.com/grunwaldlab/Population_Genetics_in_R/blob/master/apstest.txt).
If your results are different, and you don't understand why, send us an email with the report as an attachment.


7. **If you have never worked with R, please review the [Introduction to R](intro_to_R.html) page**


## Your first homework


Some aspects of analyzing genetic data are rather technical.
Others are more stylistic.
An example is the presentation of the data.
The presentation of data may include choices in color schemes and sometimes the perspective on the data.
Below is a little example that you can copy and paste into your R console.
Explore how changing the number in `set.seed()` changes the plot.
Remember to execute the `plot_poppr_msn(partial_clone, myMsn, palette = brewer.pal(n=4, name = "Set1"))` function again.
There are other examples that are commented out (i.e., the lines begin with `#` so they are not executed).
Try removing the comment character (`#`) and see how the different parameters affect the plot.
This example should validate that you have successfully installed `poppr` and hopefully provides a fun example that may inspire you to explore more options.


```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(RColorBrewer)
library(viridisLite)
library(poppr)
data(partial_clone)
myMsn <- bruvo.msn(partial_clone, include.ties = TRUE, showplot = FALSE)
set.seed(9)
plot_poppr_msn(partial_clone, myMsn, palette = brewer.pal(n=4, name = "Set1"))
#plot_poppr_msn(partial_clone, myMsn, palette = magma(n=4, begin = 0.2, end = 0.8))
#plot_poppr_msn(partial_clone, myMsn, palette = plasma(n=4))
```


## Workshop agenda

1. Introduction    
1. Introduction to [VCF data](reading_vcf.html)    
1. Analysis of [Genotyping by sequencing](gbs_analysis.html) data
1. Analysis of [Genome](analysis_of_genome.html) data


***

## How'd we do?

After you've completed the workshop, please fill out [this quick survey](https://docs.google.com/forms/d/1Buc6fsnRxEtk0QhMxH_b3yPFhP3du83zb-4m7Hd9Xtc/edit) to provide us with feedback. Thank you!
